---
title: "Group2_18S_ConeSnailMicrobiome"
author: "AshCoello"
date: "2026-04-08"
output: html_document
---

# =========================================================

# Setup (remember to set up working directory before running) 
setwd("/Users/ashercoello/Downloads/18S")

# =========================================================

```{r setup, message=FALSE, warning=FALSE}

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

cran_packages <- c(
  "ggplot2",
  "dplyr",
  "tidyr",
  "tibble",
  "vegan",
  "rstatix",
  "ape",
  "pegas",
  "scales"
)

bioc_packages <- c(
  "phyloseq",
  "microbiome"
)

for (pkg in cran_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  library(pkg, character.only = TRUE)
}
```

# =========================================================

# Find files automatically

# =========================================================

```{r find_files}
all_files <- list.files(recursive = TRUE, full.names = TRUE)
print(all_files)

find_first_match <- function(patterns, files) {
  for (pat in patterns) {
    match <- files[grepl(pat, files, ignore.case = TRUE)]
    if (length(match) > 0) {
      return(match[1])
    }
  }
  return(NA)
}

feature_table_file <- find_first_match(
  c("18S_asv-table\\.qza$", "table\\.qza$", "feature.*qza$"),
  all_files
)

taxonomy_file <- find_first_match(
  c("taxonomy\\.qza$", "taxonomy.*\\.qza$"),
  all_files
)

tree_file <- find_first_match(
  c("tree\\.qza$", "rooted.*\\.qza$", ".*\\.qza$"),
  all_files
)

metadata_file <- find_first_match(
  c("18S_Metadata\\.txt$", "metadata.*\\.txt$", "metadata.*\\.tsv$", "metadata.*\\.csv$"),
  all_files
)

shannon_file <- find_first_match(
  c("alpha-diversity\\.qza$", "shannon.*\\.qza$"),
  all_files
)

bray_file <- find_first_match(
  c("distance-matrix\\.qza$", "bray.*\\.qza$"),
  all_files
)

# This creates the "metadata_df" object the rest of the script is looking for
metadata_df <- read.table(metadata_file, header = TRUE, sep = "\t", row.names = 1)
# =========================================================

# Read the metadata

# =========================================================

```{r read_metadata}
# 1. Re-read the metadata and specify that the first column is the row name
metadata_df <- read.table(
  metadata_file, 
  header = TRUE, 
  sep = "\t", 
  row.names = 1,      # This turns the "SUH" column into the ID labels
  check.names = FALSE
)

# 2. Check to make sure the row names now look like "SUH125"
head(rownames(metadata_df))

if (!"Sample_Type" %in% colnames(metadata_df)) {
  stop("Column 'Sample_Type' not found in metadata.")
}

metadata_df$Sample_Type <- as.factor(metadata_df$Sample_Type)

print(head(metadata_df))
print(colnames(metadata_df))
print(table(metadata_df$Sample_Type))
```

# ========================================================

# Read the feature table

# =========================================================

```{r read_feature_table}
library(qiime2R)

# This skips the manual read.table headache
asv_table <- read_qza(feature_table_file)
otu_df <- asv_table$data

# Remove taxonomy column if it exists
if ("taxonomy" %in% colnames(otu_df)) {
  otu_df <- otu_df[, colnames(otu_df) != "taxonomy", drop = FALSE]
}

otu_matrix <- as.matrix(otu_df)
mode(otu_matrix) <- "numeric"

print(dim(otu_matrix))
print(otu_matrix[1:min(5, nrow(otu_matrix)), 1:min(5, ncol(otu_matrix))])
```

# ========================================================

# Read Taxonomy

# =========================================================

```{r read_taxonomy}

library(qiime2R)

# 1. Read the taxonomy artifact directly
tax_qza <- read_qza(taxonomy_file)
tax_df <- tax_qza$data

# 2. Rename columns to match the rest of your script
colnames(tax_df)[1] <- "OTU"
colnames(tax_df)[2] <- "Taxon"

# 3. Clean up the taxonomy string (Optional but recommended)
# This splits the long "k__Bacteria; p__Proteobacteria..." string into separate columns

print(colnames(tax_df))
print(head(tax_df))

# Rename first column to OTU regardless of its original name
colnames(tax_df)[1] <- "OTU"

# Make sure second column is called Taxon
if (!"Taxon" %in% colnames(tax_df)) {
  colnames(tax_df)[2] <- "Taxon"
}

tax_df_clean <- tax_df %>%
  tidyr::separate(
    col = Taxon,
    into = c(
      "Kingdom",
      "Phylum",
      "Class",
      "Order",
      "Family",
      "Genus",
      "Species"
    ),
    sep = ";",
    fill = "right",
    extra = "drop"
  ) %>%
  mutate(across(
    c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
    ~ gsub("^\\s*[a-z]__", "", .)
  )) %>%
  mutate(across(
    c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
    ~ trimws(.)
  ))

rownames(tax_df_clean) <- tax_df_clean$OTU

tax_matrix <- as.matrix(
  tax_df_clean[, c(
    "Kingdom",
    "Phylum",
    "Class",
    "Order",
    "Family",
    "Genus",
    "Species"
  )]
)

print(dim(tax_matrix))
print(head(tax_df_clean))

```

# =========================================================

# Read Tree

# =========================================================

```{r read_tree}

library(qiime2R)

# Replace line 231 with this:
tree_qza <- read_qza(tree_file)
tree <- tree_qza$data

# Now you can print it or use it in phyloseq
print(tree)

```

# ========================================================

# Match Taxa with Samples

# =========================================================

```{r match_taxa_samples}
# 1. Clean up the feature table column names
colnames(otu_matrix) <- trimws(colnames(otu_matrix))

# 2. Clean up the metadata row names
rownames(metadata_df) <- trimws(rownames(metadata_df))

# 3. Test the match again
shared_samples <- intersect(colnames(otu_matrix), rownames(metadata_df))
length(shared_samples)

shared_taxa <- intersect(rownames(otu_matrix), rownames(tax_matrix))

if (length(shared_taxa) == 0) {
  stop("No shared taxa between feature table and taxonomy.")
}

otu_matrix <- otu_matrix[shared_taxa, , drop = FALSE]
tax_matrix <- tax_matrix[shared_taxa, , drop = FALSE]

shared_samples <- intersect(colnames(otu_matrix), rownames(metadata_df))

if (length(shared_samples) == 0) {
  stop("No shared samples between feature table and metadata.")
}

otu_matrix_filtered <- otu_matrix[, shared_samples, drop = FALSE]
metadata_df_filtered <- metadata_df[shared_samples, , drop = FALSE]

cat("Number of shared taxa:", length(shared_taxa), "\n")
cat("Number of shared samples:", length(shared_samples), "\n")
```

# =========================================================

# Load phyloseq

# =========================================================

```{r load_phyloseq}

if (!requireNamespace("phyloseq", quietly = TRUE)) {
  BiocManager::install("phyloseq", ask = FALSE)
}

library(phyloseq)

```

# =========================================================

# Phyloseq

# =========================================================

```{r phyloseq}

OTU <- otu_table(otu_matrix_filtered, taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(tax_matrix))
META <- sample_data(metadata_df_filtered)
TREE <- phy_tree(tree)

physeq <- phyloseq(OTU, TAX, META, TREE)

print(physeq)

```

# =========================================================

# Alpha Diversity (Shannon)

# =========================================================

```{r alpha_diversity_data}

# --- REPLACE THE READ.TABLE BLOCK WITH THIS ---

# Use qiime2R to read the diversity artifact directly
shannon_qza <- qiime2R::read_qza(shannon_file)

# The actual data is stored in the $data object
alpha_data <- as.data.frame(shannon_qza$data)

# Rename the column to Shannon for clarity
colnames(alpha_data)[1] <- "Shannon"

alpha_data <- alpha_data[rownames(alpha_data) %in% rownames(metadata_df_filtered), , drop = FALSE]
alpha_data$Sample_Type <- metadata_df_filtered[rownames(alpha_data), "Sample_Type"]
alpha_data$Sample_ID <- rownames(alpha_data)

print(head(alpha_data))

```

```{r alpha_diversity_plot}
alpha_plot <- ggplot(alpha_data, aes(x = Sample_Type, y = Shannon, fill = Sample_Type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2, color = "black") +
  scale_fill_manual(
    values = c(
      "Water_Control_MX" = "#00BFC4",
      "Whole_venom_gland" = "#F8766D"
    )
  ) +
  labs(
    title = "Alpha Diversity (Shannon Index)",
    x = "Sample Type",
    y = "Shannon Diversity"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(alpha_plot)
```

```{r alpha_diversity_stats}

kruskal_test_result <- rstatix::kruskal_test(alpha_data, Shannon ~ Sample_Type)
print(kruskal_test_result)

```

# =========================================================

# Beta Diversity (Bray-Curtis)

# =========================================================

```{r bray_distance}
# --- REPLACE THE BRAY_DF READ.TABLE BLOCK WITH THIS ---

# 1. Use qiime2R to read the distance matrix artifact directly
bray_qza <- qiime2R::read_qza(bray_file)

# 2. Extract the data into a data frame
bray_df <- as.data.frame(as.matrix(bray_qza$data))

# 3. View the result to ensure it loaded
print(dim(bray_df))

shared_bray_samples <- intersect(rownames(bray_df), rownames(metadata_df_filtered))
bray_df <- bray_df[shared_bray_samples, shared_bray_samples, drop = FALSE]
metadata_bray <- metadata_df_filtered[shared_bray_samples, , drop = FALSE]

bray_mat <- as.matrix(bray_df)
bray_dist <- as.dist(bray_mat)

print(dim(bray_mat))
print(bray_mat[1:min(5, nrow(bray_mat)), 1:min(5, ncol(bray_mat))])

```

```{r pcoa_data}

pcoa_ord <- ape::pcoa(bray_dist)

pcoa_vectors <- as.data.frame(pcoa_ord$vectors[, 1:2])
colnames(pcoa_vectors) <- c("Axis.1", "Axis.2")
pcoa_vectors$SampleID <- rownames(pcoa_vectors)
pcoa_vectors$Sample_Type <- metadata_bray[pcoa_vectors$SampleID, "Sample_Type"]

percent_explained_1 <- round(pcoa_ord$values$Relative_eig[1] * 100, 2)
percent_explained_2 <- round(pcoa_ord$values$Relative_eig[2] * 100, 2)

print(head(pcoa_vectors))

```

```{r pcoa_plot}
pcoa_plot <- ggplot(
  pcoa_vectors,
  aes(x = Axis.1, y = Axis.2, color = Sample_Type, shape = Sample_Type)
) +
  geom_point(size = 4) +
  stat_ellipse(type = "t", level = 0.95) +
  scale_color_manual(
    values = c(
      "Water_Control_MX" = "#00BFC4",
      "Whole_venom_gland" = "#F8766D"
    )
  ) +
  labs(
    title = "PCoA of Microbial Communities",
    subtitle = "Bray-Curtis Dissimilarity",
    x = paste0("PCoA 1 (", percent_explained_1, "%)"),
    y = paste0("PCoA 2 (", percent_explained_2, "%)")
  ) +
  theme_minimal()

print(pcoa_plot)
```

```{r permanova}

sample_df <- data.frame(metadata_bray)

permanova_result <- vegan::adonis2(
  bray_dist ~ Sample_Type,
  data = sample_df,
  permutations = 999
)

print(permanova_result)

```

# =========================================================

# AMOVA

# =========================================================

```{r amova}

sample_df$Sample_Type <- as.factor(sample_df$Sample_Type)

amova_result <- pegas::amova(
  bray_dist ~ Sample_Type,
  data = sample_df,
  nperm = 999
)

print(amova_result)

```

# =========================================================

# List of Taxonomic Composition at Phylum Level

# =========================================================

```{r list_all_phyla}
all_phyla <- phyloseq::psmelt(physeq) %>%
  dplyr::mutate(
    Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", Phylum)
  ) %>%
  dplyr::distinct(Phylum) %>%
  dplyr::arrange(Phylum)

print(all_phyla)

cat("Number of unique phyla:", nrow(all_phyla), "\n")
```

# =========================================================

# Pie Chart Taxonomic Composition at Phylum Level

# =========================================================

```{r phylum_pie_chart, fig.width=10, fig.height=10}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(RColorBrewer)

phylum_pie_df <- phyloseq::psmelt(physeq) %>%
  dplyr::mutate(
    Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", Phylum)
  ) %>%
  dplyr::group_by(Phylum) %>%
  dplyr::summarise(
    TotalAbundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::arrange(desc(TotalAbundance))

# Keep only top 10 phyla
top10_phyla <- phylum_pie_df$Phylum[1:min(10, nrow(phylum_pie_df))]

phylum_pie_df$Phylum <- ifelse(
  phylum_pie_df$Phylum %in% top10_phyla,
  phylum_pie_df$Phylum,
  "Other"
)

phylum_pie_df <- phylum_pie_df %>%
  dplyr::group_by(Phylum) %>%
  dplyr::summarise(
    TotalAbundance = sum(TotalAbundance),
    .groups = "drop"
  ) %>%
  dplyr::arrange(desc(TotalAbundance)) %>%
  dplyr::mutate(
    Percentage = TotalAbundance / sum(TotalAbundance),
    Label = paste0(Phylum, " (", scales::percent(Percentage, accuracy = 0.1), ")")
  )

phylum_pie_chart <- ggplot(
  phylum_pie_df,
  aes(x = "", y = TotalAbundance, fill = Phylum)
) +
  geom_bar(
    stat = "identity",
    width = 1,
    color = "white"
  ) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Top 10 Phyla Composition",
    fill = "Phylum"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 11),
    legend.position = "right"
  )

print(phylum_pie_chart)
```

# =========================================================

# Taxonomic Composition at Phylum Level (Top 10 + Other)

# =========================================================

```{r phylum_plot, fig.width=15, fig.height=9}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(RColorBrewer)

physeq_phylum <- phyloseq::tax_glom(physeq, taxrank = "Phylum", NArm = FALSE)

tax_table(physeq_phylum)[, "Phylum"][
  is.na(tax_table(physeq_phylum)[, "Phylum"])
] <- "Unclassified"

physeq_phylum_rel <- phyloseq::transform_sample_counts(
  physeq_phylum,
  function(x) {
    if (sum(x) == 0) x else x / sum(x)
  }
)

phylum_df <- phyloseq::psmelt(physeq_phylum_rel)

phylum_df$Phylum[is.na(phylum_df$Phylum) | phylum_df$Phylum == ""] <- "Unclassified"
phylum_df <- phylum_df %>% filter(!is.na(Sample_Type))

# Keep original Sample_Type order
sample_type_levels <- unique(as.character(sample_data(physeq)$Sample_Type))
phylum_df$Sample_Type <- factor(phylum_df$Sample_Type, levels = sample_type_levels)

phylum_sums <- phylum_df %>%
  group_by(Phylum) %>%
  summarise(Total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Total))

top10_phyla <- phylum_sums$Phylum[1:min(10, nrow(phylum_sums))]

phylum_df$Phylum <- ifelse(
  phylum_df$Phylum %in% top10_phyla,
  phylum_df$Phylum,
  "Other"
)

phylum_summary <- phylum_df %>%
  group_by(Sample_Type, Phylum) %>%
  summarise(Abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

# Force every Sample_Type to appear, even if one is missing
all_phyla <- unique(phylum_summary$Phylum)

phylum_summary <- phylum_summary %>%
  tidyr::complete(
    Sample_Type = factor(sample_type_levels, levels = sample_type_levels),
    Phylum = all_phyla,
    fill = list(Abundance = 0)
  ) %>%
  group_by(Sample_Type) %>%
  mutate(TotalSample = sum(Abundance, na.rm = TRUE)) %>%
  mutate(Abundance = ifelse(TotalSample == 0, 0, Abundance / TotalSample)) %>%
  ungroup() %>%
  select(-TotalSample)

# Put Other last in legend
phylum_levels <- c(sort(setdiff(unique(phylum_summary$Phylum), "Other")), "Other")
phylum_summary$Phylum <- factor(phylum_summary$Phylum, levels = phylum_levels)

phylum_plot <- ggplot(
  phylum_summary,
  aes(x = Sample_Type, y = Abundance, fill = Phylum)
) +
  geom_bar(
    stat = "identity",
    position = "stack",
    width = 0.8
  ) +
  scale_fill_brewer(palette = "Set3") +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0, 0)
  ) +
  labs(
    title = "Relative Abundance by Phylum (Top 10 + Other)",
    x = "Sample Type",
    y = "Relative Abundance (%)",
    fill = "Phylum"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1, size = 14),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 11),
    legend.position = "bottom",
    legend.box = "vertical"
  ) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE))

print(phylum_plot)
```

# =========================================================

# Taxonomic Composition at Genus Level (Top 10 + Other)

# =========================================================

```{r genus_plot, fig.width=15, fig.height=9}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(RColorBrewer)

physeq_genus <- phyloseq::tax_glom(physeq, taxrank = "Genus", NArm = FALSE)

tax_table(physeq_genus)[, "Genus"][
  is.na(tax_table(physeq_genus)[, "Genus"])
] <- "Unclassified"

physeq_genus_rel <- phyloseq::transform_sample_counts(
  physeq_genus,
  function(x) {
    if (sum(x) == 0) x else x / sum(x)
  }
)

genus_df <- phyloseq::psmelt(physeq_genus_rel)

genus_df$Genus[is.na(genus_df$Genus) | genus_df$Genus == ""] <- "Unclassified"
genus_df <- genus_df %>% filter(!is.na(Sample_Type))

# Keep original Sample_Type order
sample_type_levels <- unique(as.character(sample_data(physeq)$Sample_Type))
genus_df$Sample_Type <- factor(genus_df$Sample_Type, levels = sample_type_levels)

genus_sums <- genus_df %>%
  group_by(Genus) %>%
  summarise(Total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Total))

top10_genera <- genus_sums$Genus[1:min(10, nrow(genus_sums))]

genus_df$Genus <- ifelse(
  genus_df$Genus %in% top10_genera,
  genus_df$Genus,
  "Other"
)

genus_summary <- genus_df %>%
  group_by(Sample_Type, Genus) %>%
  summarise(Abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

# Force every Sample_Type to appear, even if one is missing
all_genera <- unique(genus_summary$Genus)

genus_summary <- genus_summary %>%
  tidyr::complete(
    Sample_Type = factor(sample_type_levels, levels = sample_type_levels),
    Genus = all_genera,
    fill = list(Abundance = 0)
  ) %>%
  group_by(Sample_Type) %>%
  mutate(TotalSample = sum(Abundance, na.rm = TRUE)) %>%
  mutate(Abundance = ifelse(TotalSample == 0, 0, Abundance / TotalSample)) %>%
  ungroup() %>%
  select(-TotalSample)

# Put Other last in legend
genus_levels <- c(sort(setdiff(unique(genus_summary$Genus), "Other")), "Other")
genus_summary$Genus <- factor(genus_summary$Genus, levels = genus_levels)

genus_plot <- ggplot(
  genus_summary,
  aes(x = Sample_Type, y = Abundance, fill = Genus)
) +
  geom_bar(
    stat = "identity",
    position = "stack",
    width = 0.8
  ) +
  scale_fill_brewer(palette = "Set3") +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0, 0)
  ) +
  labs(
    title = "Relative Abundance by Genus (Top 10 + Other)",
    x = "Sample Type",
    y = "Relative Abundance (%)",
    fill = "Genus"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1, size = 14),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 11),
    legend.position = "bottom",
    legend.box = "vertical"
  ) +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE))

print(genus_plot)
```

# =========================================================

# Heatmap of Relative Abundance at Phylum Level

# =========================================================

```{r HeatmapPhylum1}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# Aggregate at Phylum level
physeq_heatmap <- phyloseq::tax_glom(physeq, taxrank = "Phylum", NArm = FALSE)

tax_table(physeq_heatmap)[, "Phylum"][
  is.na(tax_table(physeq_heatmap)[, "Phylum"])
] <- "Unclassified"

# Convert to relative abundance
physeq_heatmap_rel <- phyloseq::transform_sample_counts(
  physeq_heatmap,
  function(x) {
    if (sum(x) == 0) x else x / sum(x)
  }
)

heatmap_df <- phyloseq::psmelt(physeq_heatmap_rel)

heatmap_df$Phylum[
  is.na(heatmap_df$Phylum) | heatmap_df$Phylum == ""
] <- "Unclassified"

# Keep only top 10 phyla overall
top10_phyla <- heatmap_df %>%
  group_by(Phylum) %>%
  summarise(TotalAbundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TotalAbundance)) %>%
  slice(1:10) %>%
  pull(Phylum)

heatmap_df <- heatmap_df %>%
  filter(Phylum %in% top10_phyla)

# Average abundance by Sample_Type and Phylum
heatmap_summary <- heatmap_df %>%
  group_by(Sample_Type, Phylum) %>%
  summarise(
    Abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Keep sample types in original order
sample_type_levels <- unique(as.character(sample_data(physeq)$Sample_Type))
heatmap_summary$Sample_Type <- factor(
  heatmap_summary$Sample_Type,
  levels = sample_type_levels
)

# Order phyla by total abundance
phylum_order <- heatmap_summary %>%
  group_by(Phylum) %>%
  summarise(Total = sum(Abundance), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  pull(Phylum)

heatmap_summary$Phylum <- factor(
  heatmap_summary$Phylum,
  levels = rev(phylum_order)
)

phylum_heatmap <- ggplot(
  heatmap_summary,
  aes(x = Sample_Type, y = Phylum, fill = Abundance)
) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(
    low = "lightyellow",
    high = "darkred",
    labels = percent_format(accuracy = 1)
  ) +
  labs(
    title = "Heatmap of Top 10 Phyla Relative Abundance",
    x = "Sample Type",
    y = "Phylum",
    fill = "Relative\nAbundance"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 13),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

print(phylum_heatmap)
```

```{r HeatmapPhylum2}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# Aggregate counts at Phylum level
physeq_phylum_heat <- phyloseq::tax_glom(physeq, taxrank = "Phylum", NArm = FALSE)

# Replace missing phylum names
tax_table(physeq_phylum_heat)[, "Phylum"][
  is.na(tax_table(physeq_phylum_heat)[, "Phylum"]) |
  tax_table(physeq_phylum_heat)[, "Phylum"] == ""
] <- "Unclassified"

# Transform to relative abundance
physeq_phylum_heat_rel <- phyloseq::transform_sample_counts(
  physeq_phylum_heat,
  function(x) {
    if (sum(x) == 0) x else x / sum(x)
  }
)

# Convert to long format
heatmap_df <- phyloseq::psmelt(physeq_phylum_heat_rel)

heatmap_df$Phylum[
  is.na(heatmap_df$Phylum) | heatmap_df$Phylum == ""
] <- "Unclassified"

heatmap_df <- heatmap_df %>%
  filter(!is.na(Sample_Type))

# Keep Sample_Type order from metadata
sample_type_levels <- unique(as.character(sample_data(physeq)$Sample_Type))

# Average relative abundance by Sample_Type and Phylum
heatmap_summary <- heatmap_df %>%
  group_by(Sample_Type, Phylum) %>%
  summarise(
    Abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Make sure all combinations appear
all_phyla <- unique(heatmap_summary$Phylum)

heatmap_summary <- heatmap_summary %>%
  tidyr::complete(
    Sample_Type = factor(sample_type_levels, levels = sample_type_levels),
    Phylum = all_phyla,
    fill = list(Abundance = 0)
  )

# Order phyla by total abundance
phylum_order <- heatmap_summary %>%
  group_by(Phylum) %>%
  summarise(Total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  pull(Phylum)

heatmap_summary$Sample_Type <- factor(
  heatmap_summary$Sample_Type,
  levels = sample_type_levels
)

heatmap_summary$Phylum <- factor(
  heatmap_summary$Phylum,
  levels = rev(phylum_order)
)

# Heatmap
phylum_heatmap_all <- ggplot(
  heatmap_summary,
  aes(x = Sample_Type, y = Phylum, fill = Abundance)
) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient(
    low = "grey",
    high = "blue",
    labels = scales::percent_format(accuracy = 0.1)
  ) +
  coord_fixed() +
  labs(
    title = "Heatmap of Relative Abundance by Phylum",
    x = "Sample Type",
    y = "Phylum",
    fill = "Relative Abundance"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 65, hjust = 1, size = 5),
    axis.text.y = element_text(size = 5),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

print(phylum_heatmap_all)
```

```{r HeatmapPhylum3}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# =========================================================
# Aggregate at Phylum level
# =========================================================
physeq_phylum <- phyloseq::tax_glom(physeq, taxrank = "Phylum", NArm = FALSE)

tax_table(physeq_phylum)[, "Phylum"][
  is.na(tax_table(physeq_phylum)[, "Phylum"]) |
  tax_table(physeq_phylum)[, "Phylum"] == ""
] <- "Unclassified"

# =========================================================
# Convert to relative abundance per sample
# =========================================================
physeq_phylum_rel <- phyloseq::transform_sample_counts(
  physeq_phylum,
  function(x) {
    if (sum(x) == 0) x else x / sum(x)
  }
)

# =========================================================
# Melt to long format
# =========================================================
phylum_df <- phyloseq::psmelt(physeq_phylum_rel)

phylum_df$Phylum[
  is.na(phylum_df$Phylum) | phylum_df$Phylum == ""
] <- "Unclassified"

phylum_df <- phylum_df %>%
  filter(!is.na(Sample_Type))

# =========================================================
# Keep top 30 phyla by total abundance
# =========================================================
top_phyla <- phylum_df %>%
  group_by(Phylum) %>%
  summarise(Total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  slice(1:30) %>%
  pull(Phylum)

phylum_df <- phylum_df %>%
  filter(Phylum %in% top_phyla)

# =========================================================
# Sample order
# =========================================================
sample_order <- sample_data(physeq)$Sample_Type %>%
  as.character()

sample_names_ordered <- rownames(sample_data(physeq))

# If you want grouped ordering by Sample_Type:
sample_info <- data.frame(
  Sample = rownames(sample_data(physeq)),
  Sample_Type = as.character(sample_data(physeq)$Sample_Type),
  stringsAsFactors = FALSE
)

sample_info <- sample_info %>%
  arrange(Sample_Type, Sample)

sample_order <- sample_info$Sample

phylum_df$Sample <- factor(phylum_df$Sample, levels = sample_order)

# =========================================================
# Order phyla by total abundance
# =========================================================
phylum_order <- phylum_df %>%
  group_by(Phylum) %>%
  summarise(Total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  pull(Phylum)

phylum_df$Phylum <- factor(phylum_df$Phylum, levels = rev(phylum_order))

# =========================================================
# Add log-transformed abundance for better contrast
# small pseudocount avoids log(0)
# =========================================================
phylum_df <- phylum_df %>%
  mutate(LogAbundance = log10(Abundance + 1e-4))

# =========================================================
# Create positions for red separator lines between Sample_Type groups
# =========================================================
group_sizes <- sample_info %>%
  count(Sample_Type)

separator_positions <- cumsum(group_sizes$n) + 0.5
separator_positions <- separator_positions[-length(separator_positions)]

# Midpoints for group labels
group_centers <- data.frame(
  Sample_Type = group_sizes$Sample_Type,
  x = cumsum(group_sizes$n) - group_sizes$n / 2
)

# =========================================================
# Plot
# =========================================================
phylum_heatmap_samples <- ggplot(
  phylum_df,
  aes(x = Sample, y = Phylum, fill = LogAbundance)
) +
  geom_tile(color = "black", linewidth = 0.15) +
  geom_vline(
    xintercept = separator_positions,
    color = "red",
    linewidth = 0.6
  ) +
  scale_fill_gradient(
    low = "black",
    high = "#6bb6ff",
    name = "Abundance",
    breaks = log10(c(1e-4, 1e-3, 1e-2, 1e-1)),
    labels = c("0.01%", "0.1%", "1%", "10%")
  ) +
  labs(
    title = "Phylum Abundance Heatmap by Sample",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "black", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1,
      size = 7
    ),
    axis.text.y = element_text(size = 7),
    axis.title = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 9)
  )

phylum_heatmap_samples <- phylum_heatmap_samples +
  annotate(
    "text",
    x = group_centers$x,
    y = 1,
    label = group_centers$Sample_Type,
    color = "red",
    size = 4,
    fontface = "bold",
    vjust = 1.5
  )

print(phylum_heatmap_samples)
```

# =========================================================

# Heatmap of Relative Abundance at Genus Level

# =========================================================

```{r HeatmapGenus}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# Aggregate at Genus level
physeq_heatmap <- phyloseq::tax_glom(physeq, taxrank = "Genus", NArm = FALSE)

tax_table(physeq_heatmap)[, "Genus"][
  is.na(tax_table(physeq_heatmap)[, "Genus"])
] <- "Unclassified"

# Convert to relative abundance
physeq_heatmap_rel <- phyloseq::transform_sample_counts(
  physeq_heatmap,
  function(x) {
    if (sum(x) == 0) x else x / sum(x)
  }
)

heatmap_df <- phyloseq::psmelt(physeq_heatmap_rel)

heatmap_df$Genus[
  is.na(heatmap_df$Genus) | heatmap_df$Genus == ""
] <- "Unclassified"

# Keep only top 10 genera overall
top10_genera <- heatmap_df %>%
  group_by(Genus) %>%
  summarise(TotalAbundance = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TotalAbundance)) %>%
  slice(1:10) %>%
  pull(Genus)

heatmap_df <- heatmap_df %>%
  filter(Genus %in% top10_genera)

# Average abundance by Sample_Type and Genus
heatmap_summary <- heatmap_df %>%
  group_by(Sample_Type, Genus) %>%
  summarise(
    Abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Keep sample types in original order
sample_type_levels <- unique(as.character(sample_data(physeq)$Sample_Type))

heatmap_summary$Sample_Type <- factor(
  heatmap_summary$Sample_Type,
  levels = sample_type_levels
)

# Order genera by total abundance
genus_order <- heatmap_summary %>%
  group_by(Genus) %>%
  summarise(Total = sum(Abundance), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  pull(Genus)

heatmap_summary$Genus <- factor(
  heatmap_summary$Genus,
  levels = rev(genus_order)
)

genus_heatmap <- ggplot(
  heatmap_summary,
  aes(x = Sample_Type, y = Genus, fill = Abundance)
) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(
    low = "lightyellow",
    high = "darkred",
    labels = percent_format(accuracy = 1)
  ) +
  labs(
    title = "Heatmap of Top 10 Genera Relative Abundance",
    x = "Sample Type",
    y = "Genus",
    fill = "Relative\nAbundance"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 35,
      hjust = 1,
      vjust = 1,
      size = 10,
      margin = margin(t = 8)
    ),
    axis.text.y = element_text(size = 9),
    axis.title.x = element_text(
      size = 15,
      margin = margin(t = 25)
    ),
    axis.title.y = element_text(
      size = 15,
      margin = margin(r = 20)
    ),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 40, 20)
  )

print(genus_heatmap)
```

```{r HeatmapGenus2}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# Aggregate counts at Genus level
physeq_genus_heat <- phyloseq::tax_glom(physeq, taxrank = "Genus", NArm = FALSE)

# Replace missing genus names
tax_table(physeq_genus_heat)[, "Genus"][
  is.na(tax_table(physeq_genus_heat)[, "Genus"]) |
  tax_table(physeq_genus_heat)[, "Genus"] == ""
] <- "Unclassified"

# Transform to relative abundance
physeq_genus_heat_rel <- phyloseq::transform_sample_counts(
  physeq_genus_heat,
  function(x) {
    if (sum(x) == 0) x else x / sum(x)
  }
)

# Convert to long format
heatmap_df <- phyloseq::psmelt(physeq_genus_heat_rel)

heatmap_df$Genus[
  is.na(heatmap_df$Genus) | heatmap_df$Genus == ""
] <- "Unclassified"

heatmap_df <- heatmap_df %>%
  filter(!is.na(Sample_Type))

# Keep Sample_Type order from metadata
sample_type_levels <- unique(as.character(sample_data(physeq)$Sample_Type))

# Keep top 30 genera by total abundance
top_genera <- heatmap_df %>%
  group_by(Genus) %>%
  summarise(Total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  slice(1:30) %>%
  pull(Genus)

heatmap_df <- heatmap_df %>%
  filter(Genus %in% top_genera)

# Average relative abundance by Sample_Type and Genus
heatmap_summary <- heatmap_df %>%
  group_by(Sample_Type, Genus) %>%
  summarise(
    Abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Make sure all combinations appear
heatmap_summary <- heatmap_summary %>%
  tidyr::complete(
    Sample_Type = factor(sample_type_levels, levels = sample_type_levels),
    Genus = top_genera,
    fill = list(Abundance = 0)
  )

# Order genera by total abundance
genus_order <- heatmap_summary %>%
  group_by(Genus) %>%
  summarise(Total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  pull(Genus)

heatmap_summary$Sample_Type <- factor(
  heatmap_summary$Sample_Type,
  levels = sample_type_levels
)

heatmap_summary$Genus <- factor(
  heatmap_summary$Genus,
  levels = rev(genus_order)
)

# Heatmap
genus_heatmap_clean <- ggplot(
  heatmap_summary,
  aes(x = Sample_Type, y = Genus, fill = Abundance)
) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient(
  low = "grey90",
  high = "blue",
  limits = c(0, 0.10),
  breaks = c(0, 0.025, 0.05, 0.075, 0.10),
  labels = c("0%", "2.5%", "5%", "7.5%", "10%")
) +
  labs(
    title = "Heatmap of Relative Abundance by Genus (Top 30)",
    x = "Sample Type",
    y = "Genus",
    fill = "Relative Abundance"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 35,
      hjust = 1,
      vjust = 1,
      size = 10,
      margin = margin(t = 8)
    ),
    axis.text.y = element_text(size = 5),
    axis.title.x = element_text(size = 15, margin = margin(t = 25)),
    axis.title.y = element_text(size = 15, margin = margin(r = 20)),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 40, 20)
  )

print(genus_heatmap_clean)
```

```{r HeatmapGenus3}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# =========================================================
# Aggregate at Genus level
# =========================================================
physeq_genus <- phyloseq::tax_glom(physeq, taxrank = "Genus", NArm = FALSE)

tax_table(physeq_genus)[, "Genus"][
  is.na(tax_table(physeq_genus)[, "Genus"]) |
  tax_table(physeq_genus)[, "Genus"] == ""
] <- "Unclassified"

# =========================================================
# Convert to relative abundance per sample
# =========================================================
physeq_genus_rel <- phyloseq::transform_sample_counts(
  physeq_genus,
  function(x) {
    if (sum(x) == 0) x else x / sum(x)
  }
)

# =========================================================
# Melt to long format
# =========================================================
genus_df <- phyloseq::psmelt(physeq_genus_rel)

genus_df$Genus[
  is.na(genus_df$Genus) | genus_df$Genus == ""
] <- "Unclassified"

genus_df <- genus_df %>%
  filter(!is.na(Sample_Type))

# =========================================================
# Keep top 30 genera by total abundance
# =========================================================
top_genera <- genus_df %>%
  group_by(Genus) %>%
  summarise(Total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  slice(1:30) %>%
  pull(Genus)

genus_df <- genus_df %>%
  filter(Genus %in% top_genera)

# =========================================================
# Sample order
# =========================================================
sample_order <- sample_data(physeq)$Sample_Type %>%
  as.character()

sample_names_ordered <- rownames(sample_data(physeq))

# If you want grouped ordering by Sample_Type:
sample_info <- data.frame(
  Sample = rownames(sample_data(physeq)),
  Sample_Type = as.character(sample_data(physeq)$Sample_Type),
  stringsAsFactors = FALSE
)

sample_info <- sample_info %>%
  arrange(Sample_Type, Sample)

sample_order <- sample_info$Sample

genus_df$Sample <- factor(genus_df$Sample, levels = sample_order)

# =========================================================
# Order genera by total abundance
# =========================================================
genus_order <- genus_df %>%
  group_by(Genus) %>%
  summarise(Total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  pull(Genus)

genus_df$Genus <- factor(genus_df$Genus, levels = rev(genus_order))

# =========================================================
# Add log-transformed abundance for better contrast
# small pseudocount avoids log(0)
# =========================================================
genus_df <- genus_df %>%
  mutate(LogAbundance = log10(Abundance + 1e-4))

# =========================================================
# Create positions for red separator lines between Sample_Type groups
# =========================================================
group_sizes <- sample_info %>%
  count(Sample_Type)

separator_positions <- cumsum(group_sizes$n) + 0.5
separator_positions <- separator_positions[-length(separator_positions)]

# Midpoints for group labels
group_centers <- data.frame(
  Sample_Type = group_sizes$Sample_Type,
  x = cumsum(group_sizes$n) - group_sizes$n / 2
)

# =========================================================
# Plot
# =========================================================
genus_heatmap_samples <- ggplot(
  genus_df,
  aes(x = Sample, y = Genus, fill = LogAbundance)
) +
  geom_tile(color = "black", linewidth = 0.15) +
  geom_vline(
    xintercept = separator_positions,
    color = "red",
    linewidth = 0.6
  ) +
  scale_fill_gradient(
    low = "black",
    high = "#6bb6ff",
    name = "Abundance",
    breaks = log10(c(1e-4, 1e-3, 1e-2, 1e-1)),
    labels = c("0.01%", "0.1%", "1%", "10%")
  ) +
  labs(
    title = "Genus Abundance Heatmap by Sample",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "black", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1,
      size = 7
    ),
    axis.text.y = element_text(size = 7),
    axis.title = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 9)
  )
genus_heatmap_samples <- genus_heatmap_samples +
  annotate(
    "text",
    x = group_centers$x,
    y = 1,
    label = group_centers$Sample_Type,
    color = "red",
    size = 4,
    fontface = "bold",
    vjust = 1.5
  )
print(genus_heatmap_samples)
```
